Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-12-02 tmm.tex
Update some figures
Reduce all figures to 90%
Unify phrase usage and add comment on the optimality of the greedy solution

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with best online video watching experience, an adaptive video streaming system which can adjust video quality according to actual bandwidth condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. Then on that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm can significantly reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with an average PSNR gain of 0.83dB. The system is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

In video streaming, the network condition significantly influences the user's watching experience. To provide better network condition, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of OpenFlow-controlled network orchestration for video manycast \cite{Xue15}. Although these studies have offered a promising alternative, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

\begin{figure}[h]
\centering
\includegraphics[width = 1.0\linewidth]{Bitstream-Extraction.jpg}
\caption{Bitstream extraction of SVC \label{fig:Bitstream-Extraction}}
\end{figure}

@@

Particularly, if the difference of $I_2$ and $I_1$ is a single enhancement packet $j$, i.e. $I_2 \setminus I_1 = \{j\}$, then according to (\ref{eq:minus2}), the difference of the two reconstruction vectors can be written as
\begin{equation}
\label{eq:error_j}
e_j = ({\rm\bf I - M})^{-1} \cdot {\rm\bf T}_j \cdot c_j \: .
\end{equation}

@@

\begin{figure}[h]
\centering
\includegraphics[width = 1.0\linewidth]{GOP-Structure.png}
\caption{Group of pictures (GOP) in SVC \label{fig:GOP_Structure}}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width = 1.0\linewidth]{ModelVerification.png}
\caption{Comparison of actual and estimated distortion for sequence {\em Foreman} \label{fig:model_verification}}
\end{figure}

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Similar to the greedy \cite{GreedyAlgo} solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

@@

For long video sequences, an optimization window containing a certain number of frames is usually needed, and the priority assignment and bitstream extraction is conducted independently in each window. Suppose the size of the optimization window is $K$, i.e., there are $K$ frames involved in the above process, and each frame contains $L_Q$ quality layers (one base layer and $L_Q-1$ enhancement layers). Then the total number of enhancement packets that can be discarded is $K \cdot (L_Q-1)$, and in every greedy step, there are at most $K$ packets, i.e., the packets currently at the top of every frame (see analysis in \ref{subsec:analysis-extraction}), that can be selected to discard. Given that, the priority assignment process in each optimization window can be described in Algorithm \ref{algo:greedy}, with some terms in it explained below. The initial value $e_{full}$ is the error between the video sequence decoded from full enhancement packets and the original video sequence, already introduced in (\ref{eq:subset_error}). $\Phi_m$ is defined as:
\begin{equation}
\label{eq:R-D_impact_m}
\Phi_m = \min_{i \in I_{top}} \Phi_i \: ,
\end{equation}
where $I_{top}$ stands for the top-layer packets, those currently able to be discarded, and $\Phi_i$ denotes the RD impact of packet $i$. $\Phi_i$ is calculated based on (\ref{eq:rd_impact}):
\begin{equation}
\label{eq:R-D_impact_i}
\Phi_i = \dfrac{\partial D}{\partial R} = \dfrac{PSNR(e_{seq}) - PSNR(e_{seq} + e_i)}{SIZE(i)} \:,
\end{equation}
where $SIZE(i)$ is the size of packet $i$, and $PSNR(*)$ is the function to obtain PSNR from an error vector. To be clear, PSNR is defined via mean squared error (MSE), i.e., the average of the squares of the elements in the error vector.

\begin{algorithm}
\caption{Greedy-like bitstream extraction algorithm}
\label{algo:greedy}
\begin{algorithmic}
    \STATE Initialize $e_{seq}$ to $e_{full}$, and $p$ to 0.
    \WHILE{$p < K \cdot (L_Q - 1)$}
	    \FOR{the $K$ top layer packets}
		    \STATE Find the packet $m$ which has least RD impact $\Phi_m$
		\ENDFOR
		\STATE Set the priority value of packet $m$ to $p$
		\STATE Remove $m$ from the sequence
		\STATE Update $e_{seq}$ to $e_{seq} + e_m$, and update $p$ to $p+1$
    \ENDWHILE
\end{algorithmic}
\end{algorithm}

It is easy to see that the computational complexity of this priority assignment algorithm is $O(K^2)$. Note that this computational complexity is negligible compared with the decoding process needed in the error vector acquisition. With a larger optimization window we have more packets to compare and determine their priorities, which will potentially result in more optimized extraction. However, doing this will also increase the computational complexity and, more noticeably, the memory used to store the error vectors. Besides, if the priority values are to be stored and signaled using the 6-bit priority identifier field of the NALU header \cite{Amonou07}, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

After each packet is assigned a priority value and optimized extraction is made possible, the video stream now can well adapt to different data rates. Next, we propose the quality control algorithm based on the PID mechanism, which is responsible for determining the suitable data rate for transmission according to current network condition.

@@

The video quality is a continuous value. In practical operation, it is not possible to control it at an arbitrary precision. So we propose a \textit{quality level} concept as the discrete description of the video quality. When the video quality needs to be adjusted, we simply increase or decrease the quality level. Each quality level corresponds to a certain bitrate. A higher quality level corresponds to a higher bitrate and vice versa. The number of quality levels and how they are mapped to sub-streams with different bitrates depends on actual requirement or implementation. The more quality levels are defined, the more precisely the control algorithm can adjust bitrate to follow the bandwidth change. As pointed out in section \ref{sec:intro}, simulcast streaming systems can only have limited quality levels while SVC streaming systems can define quality levels at a much finer granularity.

@@

In DSS, two timelines are important. Push timeline indicates the time of pushing video data packets to streaming buffer, and play timeline indicates the time these data packet should be played, i.e., the presentation time stamp (often seen as PTS) of each packet. Their relationship is shown in Fig. \ref{fig:intervals}. The program will check the timelines periodically, and, in the default quality control algorithm of DSS, use the packet delays to decide whether quality change is needed. The two intervals shown in Fig. \ref{fig:intervals}, i.e., check interval and actual interval, are worth notice and may provide useful information. In the ideal situation, during each check interval, the server pushes an amount of data that will actually play for a period of time equal to that check interval, i.e., \textit{actual\_interval} = \textit{check\_interval}. If \textit{actual\_interval} is smaller than \textit{check\_interval}, it means that less data is pushed to the client, and the pushing is slowed down because bandwidth is relatively low at this time. On the other hand, if \textit{actual\_interval} is larger than \textit{check\_interval}, it indicates the bandwidth condition is good and data is being pushed quickly to the client. The ratio of these two intervals, i.e., \textit{actual\_check interval ratio}, is in direct proportion to the ratio of the data rate appropriate for bandwidth and the current pushing data rate.

The \textit{actual\_check interval ratio} is suitable for quality control because: 1) this ratio directly reflects the mismatch of current pushing data rate and the bandwidth, and by adjusting data rate according to it, this ratio can be used to accurately change video quality; 2) the desired value of this ratio is 1.0 by definition so we don't need to find the optimized setpoint by mathematical methods or experiments; and 3) it is easy to compute and implement in practical systems.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{Intervals.jpg}
\caption{Two timelines in Darwin Streaming Sever (DSS) \label{fig:intervals}}
\end{figure}

@@

As is introduced above, the \textit{actual\_check interval ratio} serves as the process variable in this PID-based quality control algorithm. In normal PID control, the controller output is calculated using the proportion, integral and differential of the error between the process variable and the setpoint, as shown in (\ref{eq:pid-output}). Considering that the chosen process variable is a ratio and the corresponding setpoint is 1, we make a little modification to the general PID controller and obtain a ratio-based model. In this model, the ratio is directly used to calculate the controller output, and all subtraction operations in original PID equations have been changed to division operations to keep the correctness, as follows:

@@

In (\ref{eq:ut}), $K_p$, $K_i$, and $K_d$ indicate the tuning parameters for proportional, integral and derivative parts respectively, and $E_p^t$, $E_i^t$, $E_d^t$ stand for the corresponding observed value for each part, which are defined in (\ref{eq:ep})-(\ref{eq:long-check}); the controller output $u_t$ is utilized to calculate an appropriate data rate and further to determine which quality level we should transmit (see Algorithm \ref{algo:control} for more details).

In (\ref{eq:long-actual}) and (\ref{eq:long-check}), $T_l$ stands for the last quality change time. Theoretically, the integral part should be recorded and calculated since the system starts. In practice, however, bandwidth may vary significantly during the transmission. If the integral part was recorded for too long time, it may not reflect the real network status and may slow down the reaction for better quality level. A compromise here is to choose a relatively long time, i.e., the time from last quality change till now, for the integral part calculation.

This ratio-based model is more simple and effective for the chosen process variable and setpoint. Moreover, in this model the controller output has intuitive meaning when the three parameters are normalized. It is easy to know that all of the three PID parts ($E_p^t$, $E_i^t$, $E_d^t$) are ratios with a value more or less than 1, and the controller output $u_t$ is a weighted average of the three ratios, with the three parameters as weights. Therefore the controller output is also a ratio value around 1, which reflects the relationship between the bitrate appropriate for bandwidth condition and the bitrate currently being used. In other words, if the output is larger than 1, then the controller has determined the bitrate should be increased accordingly, and vice versa. Use it as a multiplier to manipulate the bitrate currently being used, and we would be able to control the streaming system to match bandwidth condition.

The whole process of the proposed algorithm is described in Algorithm \ref{algo:control}. Compared to other scheduling strategies, this quality control algorithm has several advantages. First, this algorithm does not just rely on current packet delay or information at fixed time point, therefore quality change may occur at any proper time. Second, since the model has taken prediction into account, it is able to decide whether the bandwidth will be high or low enough for the quality level to change in the near future, thus avoiding video quality fluctuation. Third, in this algorithm it's not necessary to change one quality level at a time. This enables fast quality drop to avoid a latent playback pause when bandwidth suddenly decreases, and fast quality rise to provide better video as soon as possible when system starts, resulting in a seamless watching experience where the user can enjoy the video without stopping or waiting.

@@

In this work we simply use the standard Ziegler--Nichols method which already takes above impacts into consideration and produces acceptably good results. Adopting more sophisticated or automated techniques to further optimize the parameters will be done in the future. It should also be noted that, in the video quality control algorithm only the data rate and the bandwidth are relevant, not the video content. Therefore once the parameters are chosen they should apply well to different video streams.

@@

\begin{figure*}[t]
\centering
\subfloat[City]{
\includegraphics[width=0.5\textwidth]{City.jpg}
\label{fig:City}}
\subfloat[Foreman]{
\includegraphics[width=0.5\textwidth]{Foreman.jpg}
\label{fig:Foreman}}
\qquad
\subfloat[Harbour]{
\includegraphics[width=0.5\textwidth]{Harbour.jpg}
\label{fig:Harbour}}
\subfloat[Soccer]{
\includegraphics[width=0.5\textwidth]{Soccer.jpg}
\label{fig:Soccer}}
\caption{Performance of three bitstream extraction methods for various CIF sequences}
\label{fig:extraction-performance}
\end{figure*}

@@

We simulate network fluctuation by adjusting bandwidth around a certain level. We set this level from 400 to 1400 kbps to reflect different average bandwidth. The simulation is implemented with NetLimiter 3 \cite{Netlimiter}, where the bandwidth varies randomly with certain distribution and intense fluctuations are observed. The baseline algorithm in the comparison refers to the packet delay feedback (PDF) algorithm in the original DSS.

@@

\begin{figure*}[t]
\subfloat[Average quality level of different control algorithms]{
\includegraphics[width=0.5\textwidth]{Quality-all.png}
\label{fig:quality-all}}
\subfloat[Quality level variance of different control algorithms]{
\includegraphics[width=0.5\textwidth]{Variance-all.png}
\label{fig:variance-all}}
\caption{Comparison of the long/short time prediction and the PDF algorithm in DSS}
\label{fig:performance-all}
\end{figure*}

\begin{figure*}[t]
\centering
\subfloat[Average quality level of PID and PDF control algorithms]{
\includegraphics[width=0.5\textwidth]{Quality-two.png}
\label{fig:quality-two}}
\subfloat[Quality level variance of PID and PDF control algorithms]{
\includegraphics[width=0.5\textwidth]{Variance-two.png}
\label{fig:variance-two}}
\caption{Comparison of the proposed PID algorithm and the PDF algorithm in DDS}
\label{fig:performance-two}
\end{figure*}

@@

\begin{figure*}[t]
\centering
\subfloat[Bandwidth fluctuates around 400kbps: PSNR average and variance in the original system are 36.00 and 0.30, while in the proposed system are 36.50 (increased by 0.50) and 0.23 (reduced by 0.07).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR1.png}
\label{fig:Fluctuation-PSNR1}}\hfill
\subfloat[Bandwidth fluctuates around 600kbps: PSNR average and variance in the original system are 37.06 and 0.61, while in the proposed system are 37.32 (increased by 0.26) and 0.33 (reduced by 0.28).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR2.png}
\label{fig:Fluctuation-PSNR2}}
\qquad
\subfloat[Bandwidth fluctuates around 800kbps: PSNR average and variance in the original system are 37.92 and 1.24, while in the proposed system are 38.75 (increased by 0.83) and 0.69 (reduced by 0.55).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR3.png}
\label{fig:Fluctuation-PSNR3}}\hfill
\subfloat[Bandwidth fluctuates around 1000kbps: PSNR average and variance in the original system are 38.28 and 1.30, while in the proposed system are 38.94 (increased by 0.66) and 0.71 (reduced by 0.59).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR4.png}
\label{fig:Fluctuation-PSNR4}}
\caption{The streamed video's PSNR change with time when bandwidth fluctuates
\label{fig:fluctuation-psnr}}
\end{figure*}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with best online video watching experience, an adaptive video streaming system which can adjust video quality according to actual network condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. Then on that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm can significantly reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with an average PSNR gain of 0.83dB. The system is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

In video streaming, the network condition significantly influences the user's watching experience. To provide better network condition, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of OpenFlow-controlled network orchestration for video manycast \cite{Xue15}. Although these studies have offered a promising alternative, currently traditional best-effort networks still dominate, and the variety of users' network condition remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the network condition. In other words, the video streaming system needs to be adaptive.

@@

\begin{figure}[h]
\centering
\includegraphics[width = 0.9\linewidth]{Bitstream-Extraction.jpg}
\caption{Bitstream extraction of SVC \label{fig:Bitstream-Extraction}}
\end{figure}

@@

Particularly, if the difference of $I_2$ and $I_1$ is a single enhancement packet $j$, i.e., $I_2 \setminus I_1 = \{j\}$, then according to (\ref{eq:minus2}), the difference of the two reconstruction vectors can be written as
\begin{equation}
\label{eq:error_j}
e_j = ({\rm\bf I - M})^{-1} \cdot {\rm\bf T}_j \cdot c_j \: .
\end{equation}

@@

\begin{figure}[h]
\centering
\includegraphics[width = 0.9\linewidth]{GOP-Structure.png}
\caption{Group of pictures (GOP) in SVC \label{fig:GOP_Structure}}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{ModelVerification.png}
\caption{Comparison of actual and estimated distortion for sequence {\em Foreman} \label{fig:model_verification}}
\end{figure}

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Since finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction algorithm uses a greedy method \cite{GreedyAlgo} to get a sub-optimal solution. Note that the greedy algorithm is able to find a local optimum. However, the gap between the local optimum and the global optimum is uncertain. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

@@

For long video sequences, an optimization window containing a certain number of frames is usually needed, and the priority assignment and bitstream extraction is conducted independently in each window. Suppose the size of the optimization window is $K$, i.e., there are $K$ frames involved in the above process, and each frame contains $L_Q$ quality layers (one base layer and $L_Q-1$ enhancement layers). Then the total number of enhancement packets that can be discarded is $K \cdot (L_Q-1)$, and in every greedy step, there are at most $K$ packets, i.e., the packets currently at the top of every frame (see analysis in \ref{subsec:analysis-extraction}), that can be selected to discard. Given that, the priority assignment process in each optimization window can be described in Algorithm \ref{algo:greedy}, with some terms in it explained below. The initial value $e_{full}$ is the error between the video sequence decoded from full enhancement packets and the original video sequence, as already introduced in (\ref{eq:subset_error}). $\Phi_m$ is defined as:
\begin{equation}
\label{eq:R-D_impact_m}
\Phi_m = \min_{i \in I_{top}} \Phi_i \: ,
\end{equation}
where $I_{top}$ stands for the top-layer packets, those currently able to be discarded, and $\Phi_i$ denotes the RD impact of packet $i$. $\Phi_i$ is calculated based on (\ref{eq:rd_impact}):
\begin{equation}
\label{eq:R-D_impact_i}
\Phi_i = \dfrac{\partial D}{\partial R} = \dfrac{PSNR(e_{seq}) - PSNR(e_{seq} + e_i)}{SIZE(i)} \:,
\end{equation}
where $SIZE(i)$ is the size of packet $i$, and $PSNR(*)$ is the function to obtain PSNR from an error vector. To be clear, PSNR is defined via mean squared error (MSE), i.e., the average of the squares of the elements in the error vector.

\begin{algorithm}
\caption{Greedy-like priority assignment algorithm}
\label{algo:greedy}
\begin{algorithmic}
    \STATE Initialize $e_{seq}$ to $e_{full}$, and $q$ to 0.
    \WHILE{$q < K \cdot (L_Q - 1)$}
	    \FOR{the (at most) $K$ top layer packets}
		    \STATE Find the packet $m$ which has the least RD impact $\Phi_m$
		\ENDFOR
		\STATE Set the priority value of packet $m$ to $q$
		\STATE Remove $m$ from the sequence
		\STATE Update $e_{seq}$ to $e_{seq} + e_m$, and update $q$ to $q+1$
    \ENDWHILE
\end{algorithmic}
\end{algorithm}

It is easy to see that the computational complexity of this priority assignment algorithm is $O(K^2)$. Note that, however, the computation in the priority assignment process is actually very insignificant compared with the decoding process needed in the error vector acquisition. A larger optimization window means we have more packets to compare and determine their priorities, which will potentially result in more optimized extraction. However, doing this will also increase the computational complexity and, more noticeably, increase the memory used to store the error vectors. Besides, if the priority values are to be stored and signalled using the 6-bit priority identifier field of the NALU header \cite{Amonou07}, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

After each packet is assigned a priority value and optimized extraction is made possible, the video stream now can well adapt to different bitrates. Next, we propose the quality control algorithm based on the PID mechanism, which is responsible for determining the suitable bitrate for transmission according to current network condition.

@@

The video quality is a continuous value, and it is impractical to control it at an arbitrary precision. Therefore we propose a \textit{quality level} concept as the discrete description of the video quality. When the video quality needs to be adjusted, we simply increase or decrease the quality level. Each quality level corresponds to a certain bitrate. A higher quality level corresponds to a higher bitrate and vice versa. The number of quality levels and how they are mapped to sub-streams with different bitrates depends on actual requirement or implementation. The more quality levels are defined, the more precisely the control algorithm can adjust bitrate to follow the bandwidth change. As pointed out in section \ref{sec:intro}, simulcast streaming systems can only have limited quality levels while SVC streaming systems can define quality levels at a much finer granularity.

@@

In DSS, two timelines are important. Push timeline indicates the time of pushing video data packets to streaming buffer, and play timeline indicates the time these data packet should be played, i.e., the presentation time stamp (often seen as PTS) of each packet. Their relationship is shown in Fig. \ref{fig:intervals}. The program will check the timelines periodically, and, in the default quality control algorithm of DSS, use the packet delays to decide whether quality change is needed. The two intervals shown in Fig. \ref{fig:intervals}, i.e., check interval and actual interval, are worth notice and may provide useful information. In the ideal situation, during each check interval, the server pushes an amount of data that will actually play for a period of time equal to that check interval, i.e., \textit{actual\_interval} = \textit{check\_interval}. If \textit{actual\_interval} is smaller than \textit{check\_interval}, it means that less data is pushed to the client, and the pushing is slowed down because bandwidth is relatively low at this time. On the other hand, if \textit{actual\_interval} is larger than \textit{check\_interval}, it indicates the network condition is good and data is being pushed quickly to the client. The ratio of these two intervals, i.e., \textit{actual\_check interval ratio}, is in direct proportion to the ratio of the bitrate suitable for the network condition and the bitrate currently being used.

The \textit{actual\_check interval ratio} is suitable to use in quality control because: 1) this ratio directly reflects the mismatch of current pushing bitrate and the bandwidth, and by adjusting bitrate according to it, this ratio can be used to accurately change video quality; 2) the desired value of this ratio is 1.0 by definition so we don't need to find the optimized setpoint by mathematical methods or experiments; and 3) it is easy to compute and implement in practical systems.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{Intervals.png}
\caption{Two timelines in Darwin Streaming Sever (DSS) \label{fig:intervals}}
\end{figure}

@@

As is introduced above, the \textit{actual\_check interval ratio} serves as the process variable in this PID-based quality control algorithm. In normal PID control, the controller output is calculated using the proportion, integral and differential of the error between the process variable and the setpoint, as shown in (\ref{eq:pid-output}). Considering that the chosen process variable is a ratio and the corresponding setpoint is 1, we make a little modification to the general PID controller and obtain a ratio-based model. In this model, the ratio is directly used to calculate the controller output, and all subtraction operations in original PID equations have been properly changed to division operations to keep the correctness, as follows:

@@

In (\ref{eq:ut}), $K_p$, $K_i$, and $K_d$ indicate the tuning parameters for proportional, integral and derivative parts respectively, and $E_p^t$, $E_i^t$, $E_d^t$ stand for the corresponding observed value for each part, which are defined in (\ref{eq:ep})-(\ref{eq:long-check}); the controller output $u_t$ is utilized to calculate a suitable bitrate for transmission and further to determine which quality level should be selected (see Algorithm \ref{algo:control} for more details).

In (\ref{eq:long-actual}) and (\ref{eq:long-check}), $T_l$ stands for the last quality change time. Theoretically, the integral part should be recorded and calculated since the system starts. In practice, however, bandwidth may change significantly during the transmission. If the integral part is recorded for too long time, it may fail to reflect the current network condition and thus slow down the reaction for better quality level. A compromise here is to choose a relatively long time, i.e., the time from last quality change till now, for the integral part calculation.

This ratio-based model is more simple and effective for the chosen process variable and setpoint. Moreover, in this model the controller output has intuitive meaning when the three parameters are normalized. It is easy to know that all of the three PID parts ($E_p^t$, $E_i^t$, $E_d^t$) are ratios with a value more or less than 1, and the controller output $u_t$ can be seen as a weighted average of the three ratios, with the three parameters as weights. Therefore the controller output is also a ratio value around 1, which reflects the relationship between the bitrate suitable for the network condition and the bitrate currently being used. In other words, if the output is larger than 1, then the controller has determined the bitrate should be increased accordingly, and vice versa. Use it as a multiplier to manipulate the bitrate currently being used, and we would be able to control the streaming system to match the network condition.

The whole process of the proposed algorithm is described in Algorithm \ref{algo:control}. Compared to other scheduling strategies, this quality control algorithm has several advantages. First, this algorithm does not just rely on current packet delay or information at a fixed time point, therefore quality change may occur at any proper time. Second, since the model has taken prediction into account, it is able to decide whether the bandwidth will be high or low enough for the quality level to change in the near future, thus to avoid unnecessary quality change. Third, in this algorithm it's not obliged to change one quality level at a time. This enables fast quality drop to avoid a latent playback pause when bandwidth suddenly decreases, and fast quality rise to provide better video as soon as possible when system starts, resulting in a seamless watching experience where the user can enjoy the video without stopping or waiting.

@@

In this work we simply use the standard Ziegler--Nichols method which already takes above impacts into consideration and produces acceptably good results. Adopting more sophisticated or automated techniques to further optimize the parameters will be done in the future. It should also be noted that, in the video quality control algorithm only the bitrate and the bandwidth are relevant, not the video content. Therefore once the parameters are chosen they should apply well to different video streams.

@@

\begin{figure*}[t]
\centering
\subfloat[City]{
\includegraphics[width=0.45\textwidth]{City.jpg}
\label{fig:City}}
\subfloat[Foreman]{
\includegraphics[width=0.45\textwidth]{Foreman.jpg}
\label{fig:Foreman}}
\qquad
\subfloat[Harbour]{
\includegraphics[width=0.45\textwidth]{Harbour.jpg}
\label{fig:Harbour}}
\subfloat[Soccer]{
\includegraphics[width=0.45\textwidth]{Soccer.jpg}
\label{fig:Soccer}}
\caption{Performance of three bitstream extraction methods for various CIF sequences}
\label{fig:extraction-performance}
\end{figure*}

@@

We simulate bandwidth fluctuation by adjusting bandwidth around a certain level. We set this level from 400 to 1400 kbps to reflect different average bandwidth. The simulation is implemented with NetLimiter 3 \cite{Netlimiter}, where the bandwidth varies randomly with certain distribution and intense fluctuations are observed. The baseline algorithm in the comparison refers to the packet delay feedback (PDF) algorithm in the original DSS.

@@

\begin{figure*}[t]
\centering
\subfloat[Average quality level of different control algorithms]{
\includegraphics[width=0.45\textwidth]{Quality-all.png}
\label{fig:quality-all}}
\subfloat[Quality level variance of different control algorithms]{
\includegraphics[width=0.45\textwidth]{Variance-all.png}
\label{fig:variance-all}}
\caption{Comparison of the long/short time prediction and the PDF algorithm in DSS}
\label{fig:performance-all}
\end{figure*}

\begin{figure*}[t]
\centering
\subfloat[Average quality level of PID and PDF control algorithms]{
\includegraphics[width=0.45\textwidth]{Quality-two.png}
\label{fig:quality-two}}
\subfloat[Quality level variance of PID and PDF control algorithms]{
\includegraphics[width=0.45\textwidth]{Variance-two.png}
\label{fig:variance-two}}
\caption{Comparison of the proposed PID algorithm and the PDF algorithm in DDS}
\label{fig:performance-two}
\end{figure*}

@@

\begin{figure*}[t]
\centering
\subfloat[Bandwidth fluctuates around 400kbps: PSNR average and variance in the original system are 36.00 and 0.30, while in the proposed system are 36.50 (increased by 0.50) and 0.23 (reduced by 0.07).]{
\includegraphics[width=0.45\textwidth]{Fluctuation-PSNR1.png}
\label{fig:Fluctuation-PSNR1}}\hfill
\subfloat[Bandwidth fluctuates around 600kbps: PSNR average and variance in the original system are 37.06 and 0.61, while in the proposed system are 37.32 (increased by 0.26) and 0.33 (reduced by 0.28).]{
\includegraphics[width=0.45\textwidth]{Fluctuation-PSNR2.png}
\label{fig:Fluctuation-PSNR2}}
\qquad
\subfloat[Bandwidth fluctuates around 800kbps: PSNR average and variance in the original system are 37.92 and 1.24, while in the proposed system are 38.75 (increased by 0.83) and 0.69 (reduced by 0.55).]{
\includegraphics[width=0.45\textwidth]{Fluctuation-PSNR3.png}
\label{fig:Fluctuation-PSNR3}}\hfill
\subfloat[Bandwidth fluctuates around 1000kbps: PSNR average and variance in the original system are 38.28 and 1.30, while in the proposed system are 38.94 (increased by 0.66) and 0.71 (reduced by 0.59).]{
\includegraphics[width=0.45\textwidth]{Fluctuation-PSNR4.png}
\label{fig:Fluctuation-PSNR4}}
\caption{The streamed video's PSNR change with time when bandwidth fluctuates
\label{fig:fluctuation-psnr}}
\end{figure*}
